### 2022-11-02 response.tex

\revtext{Based on that reviewer's criticism, I am recommending what I call a soft rejection. Address this reviewer's criticism in detail and modify the manuscript accordingly. By "address," I am considering the possibility of your presenting a compelling argument for why this issue need not be considered in this report. Your response to the reviews and the revised manuscript will be sent to the second reviewer.}

\rev{Reviewer \#1}

\revtext{The paper commences with a very clear introduction to the problems in estimating the transmission advantage of a new variant - in particular, the role of dynamical bias. This is followed by a very reasonable proposal for how the problem may be addressed. Applying these methods to a re-analysis of data on within- and between- household spread of delta and omicron variants of SARS-CoV-2 suggests that while both had similar incubation-period distributions, the mean generation interval for omicron was shorter by 0.3-0.8 days than previously estimated (0.6 days). The original analysis had also reported a difference in mean incubation periods of 3.2 days for omicron versus 4.4 days for delta; the methods used in this paper suggest that both have a mean of 4.1 days. The consequences of discriminating between generation time intervals on the estimation of transmission advantage is shown in Fig 5, although it is not clear how this pertains to dynamical bias as both the original study and this one reported a difference between the variants in this respect.}

Dynamical biases pertain to the estimation of delay distributions, and not to the estimation of transmission advantage. 
We agree that both the original study and this one report that the Omicron variant has a higher reproduction number. 
Here, what we show is that the estimates of transmission advantage depends on the assumptions about the generation-interval distributions. 
For example, as we explain in the Discussion, assuming longer generation-interval distributions can lead to further biases in the estimates of transmission advantange:

``For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the previously assumed generation-interval distributions and estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.''

@@

\revtext{The caveats of this study, and the problems of applying this method to other datasets, are very thoroughly outlined in the discussion.}

Thank you.

\revtext{Less attention is paid to the more important question of WHY the omicron variant had such a short generation time in this setting.}

We have now added text to the discussion explaining different types of generation intervals (intrinsic vs realized) and why the realized generation interval of the Omicron variant might be shorter:

``Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
The realized generation intervals represent time between actual infection events and are different from the intrinsic generation intervals, which represent the changes in the infectiousness of an infected individuals over the course of their infections \citep{champredon2015intrinsic}.

Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals of the Omicron variant.
For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} can lead to faster recovery, which in turn can also shorten the intrinsic generation interval \citep{roberts2007model}.
Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
At the population-level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Omicron variant.
At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.

Even if the Delta and Omicron variants were to have identical intrinsic generation-interval distributions, their realized generation-interval distributions can still differ.
For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, can also lead to shorter realized generation intervals.
However, this explanation is also less likely given similarities in the inferred incubation-period distributions.

Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections. 
To better illustrate the network effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days;
in this case, both variants have identical intrinsic generation-interval distributions.
However, if both individuals were each living in a household consisting of four other people, an individual infected with variant A will infect one household member everyday; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
Previous simulations showed that such network effects can have considerable impact on realized generation intervals even during the initial exponential growth phase, when susceptible depletion is negligible at the population level \citep{park2020inferring}.
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).

Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but it is still uncertain how infectiousness profiles differ intrinsically between Omicron and Delta.
In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
and can therefore be expected to improve estimates of effective reproduction numbers.''

@@

\revtext{The authors discuss the possibility of a "network effect" which is predicated on omicron having a higher Ro and therefore somewhat circular.}

This explanation is not circular. As we explain in the Discussion, the success of the Omicron variant at the population-level is due to immune evasion (also pointed out by reviwer 1), and this doesn't require shorter generation interval. Instead, we now provide a concrete example for the network effect, explaining how higher reproduction number can lead to shorter \emph{realized} generation intervals even when two variants have identical \emph{intrinsic} generation intervals:

``To better illustrate the network effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days;
in this case, both variants have identical intrinsic generation-interval distributions.
However, if both individuals were each living in a household consisting of four other people, an individual infected with variant A will infect one household member everyday; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.''

@@

\revtext{Other factors "such as more stringent contact tracing measures against the Omicron variant in the Netherlands [4], faster within-host clearance of the Omicron variant [35], and viral kinetics of reinfection and breakthrough infections" are mentioned without sufficient explanation as to why they would yield such a result.}

We now provide a more detailed explanation.

\revtext{The role of immune evasion - although mentioned in passing - has not been sufficiently accommodated either in the discussion or in the analysis itself, even though it presents itself as the most likely explanation for the success of omicron in almost every global setting.}

We now discuss the possible effects of immune evasion more thorougly. While we agree that the immune evasion is the most likely explanation for the success of omicron in almost every global setting, its success or immune evasion does not require or depend on changes in generation-interval distributions.

@@

\revtext{To address this properly, the authors will have to account for the different proportions of the population currently immune to infection by either variant and how the level of cross-immunity between them affects our observations of transmission advantage. The authors should at least justify why they have not taken such an obvious feature of this multi-strain system into account.}

As explained previously, the population-level success of the Omicron variant is almost a separate question from the estimation of their incubation-period and generation-interval distributions, which is the main goal of our analysis. We also now provide a more thorough explanation for how pre-existing immunity and immune evasiveness of the Omicron variant might shape the generation-interval distribution:

``Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
At the population-level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Omicron variant.
At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.''

### 2022-11-15 response.tex

\revtext{Based on that reviewer's criticism, I am recommending what I call a soft rejection. Address this reviewer's criticism in detail and modify the manuscript accordingly. By "address," I am considering the possibility of your presenting a compelling argument for why this issue need not be considered in this report. Your response to the reviews and the revised manuscript will be sent to the second reviewer.}

\rev{Reviewer \#1}

\revtext{The paper commences with a very clear introduction to the problems in estimating the transmission advantage of a new variant - in particular, the role of dynamical bias. This is followed by a very reasonable proposal for how the problem may be addressed. Applying these methods to a re-analysis of data on within- and between- household spread of delta and omicron variants of SARS-CoV-2 suggests that while both had similar incubation-period distributions, the mean generation interval for omicron was shorter by 0.3-0.8 days than previously estimated (0.6 days). The original analysis had also reported a difference in mean incubation periods of 3.2 days for omicron versus 4.4 days for delta; the methods used in this paper suggest that both have a mean of 4.1 days. The consequences of discriminating between generation time intervals on the estimation of transmission advantage is shown in Fig 5, although it is not clear how this pertains to dynamical bias as both the original study and this one reported a difference between the variants in this respect.}

Dynamical biases pertain to the estimation of delay distributions, and not to the estimation of transmission advantage. 
We agree that both the original study and this one report that the Omicron variant has a higher reproduction number. 
Here, what we show is that estimates of transmission advantage depend on assumptions about the generation-interval distributions. 
For example, as we explain in the Discussion, assuming longer generation-interval distributions can lead to further biases in the estimates of transmission advantange:

``For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to the assumptions and results of \cite{pearson2021bounding} for Gauteng province.''

### 2022-11-16 response.tex

\revtext{Based on that reviewer's criticism, I am recommending what I call a soft rejection. Address this reviewer's criticism in detail and modify the manuscript accordingly. By "address," I am considering the possibility of your presenting a compelling argument for why this issue need not be considered in this report. Your response to the reviews and the revised manuscript will be sent to the second reviewer.}

\rev{Reviewer \#1}

\revtext{The paper commences with a very clear introduction to the problems in estimating the transmission advantage of a new variant - in particular, the role of dynamical bias. This is followed by a very reasonable proposal for how the problem may be addressed. Applying these methods to a re-analysis of data on within- and between- household spread of delta and omicron variants of SARS-CoV-2 suggests that while both had similar incubation-period distributions, the mean generation interval for omicron was shorter by 0.3-0.8 days than previously estimated (0.6 days). The original analysis had also reported a difference in mean incubation periods of 3.2 days for omicron versus 4.4 days for delta; the methods used in this paper suggest that both have a mean of 4.1 days. The consequences of discriminating between generation time intervals on the estimation of transmission advantage is shown in Fig 5, although it is not clear how this pertains to dynamical bias as both the original study and this one reported a difference between the variants in this respect.}

Dynamical biases pertain to the estimation of delay distributions, and not to the estimation of transmission advantage. 
We agree that both the original study and this one report that the Omicron variant has a higher reproduction number. 
Here, what we show is that estimates of transmission advantage depend on assumptions about the generation-interval distributions. 
For example, as we explain in the Discussion, assuming longer generation-interval distributions can lead to further biases in the estimates of transmission advantange:

``For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province.''

@@

\revtext{The caveats of this study, and the problems of applying this method to other datasets, are very thoroughly outlined in the discussion.}

Thank you.

\revtext{Less attention is paid to the more important question of WHY the omicron variant had such a short generation time in this setting.}

We have now added text to the discussion explaining different types of generation intervals (intrinsic vs realized) and why the realized generation interval of the Omicron variant might be shorter:

``Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
\jd{“tried to” is odd; can we drop it?}
\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
The realized generation intervals represent time between actual infection events, and are different from the intrinsic generation intervals, which reflect the average profile of infectiousness of infected individuals over the course of their infections \citep{champredon2015intrinsic}.

Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals.
For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} may lead to faster recovery, which could in turn shorten the intrinsic generation interval \citep{roberts2007model}.
\jd{Something about: A greater proportion of Omicron than Delta infections will be in people with pre-existing immunity (because these people are less likely to be infected by Delta).}
Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
At the population-level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
\jd{Don't understand the following; we're rebutting a point that hasn't been clearly outlined. The whole ¶ I guess could be edited to combine points above and below this comment.}
however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Delta variant.
At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.

Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions, their realized generation-interval distributions could still differ.
For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, resulting in a shorter realized generation interval.
Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, could also lead to shorter realized generation intervals.
However, this explanation is also less likely given similarities in the inferred incubation-period distributions.

Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized. 
To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days;
in this case, both variants have identical intrinsic generation-interval distributions.
We note that this scenario does not match our model, which assumes constant force of infection, not a deterministic number of infections.
However, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
\jd{It bugs me how awkwardly this matches our actual model but I think it's effective communication. I added an awkward s.~above, which you can think about.}
Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).

Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but the difference between the intrinsic infectiousness profiles remains unclear. 
In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than current estimates, given existing levels of interventions, including vaccination, and pandemic awareness.
Estimating initial intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
and can therefore be expected to improve estimates of effective reproduction numbers.''

@@

\revtext{The authors discuss the possibility of a "network effect" which is predicated on omicron having a higher Ro and therefore somewhat circular.}

We don't believe that this explanation is circular, and have tried to clarify our explication further. As we explain in the Discussion, the success of the Omicron variant at the population level is due to immune evasion (also pointed out by reviwer 1), and this doesn't require shorter generation interval. We now provide a concrete example for the network effect, explaining how higher reproduction number can lead to shorter \emph{realized} generation intervals even when two variants have identical \emph{intrinsic} generation intervals:

\jd{Reinsert this text after editing.}

@@

\revtext{Other factors "such as more stringent contact tracing measures against the Omicron variant in the Netherlands [4], faster within-host clearance of the Omicron variant [35], and viral kinetics of reinfection and breakthrough infections" are mentioned without sufficient explanation as to why they would yield such a result.}

We now provide a more detailed explanation.

\revtext{The role of immune evasion - although mentioned in passing - has not been sufficiently accommodated either in the discussion or in the analysis itself, even though it presents itself as the most likely explanation for the success of omicron in almost every global setting.}

We now discuss the possible effects of immune evasion more thorougly.

@@

\revtext{To address this properly, the authors will have to account for the different proportions of the population currently immune to infection by either variant and how the level of cross-immunity between them affects our observations of transmission advantage. The authors should at least justify why they have not taken such an obvious feature of this multi-strain system into account.}

While we agree that immune evasion is the most likely explanation for the success of omicron in almost every global setting, its success or immune evasion does not require or depend on changes in generation-interval distributions.

We now provide a more thorough explanation for how pre-existing immunity and immune evasiveness of the Omicron variant might shape the generation-interval distribution:

``Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
At the population level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Omicron variant.
At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.''

### 2022-11-28 omicron-generation.tex
done

Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
The realized generation intervals represent time between actual infection events and are different from the intrinsic generation intervals, which represent the changes in the infectiousness of an infected individuals over the course of their infections \citep{champredon2015intrinsic}.

Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals of the Omicron variant.
For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} can lead to faster recovery, which in turn can also shorten the intrinsic generation interval \citep{roberts2007model}.
Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
At the population-level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Omicron variant.
At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.

Even if the Delta and Omicron variants were to have identical intrinsic generation-interval distributions, their realized generation-interval distributions can still differ.
For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, can also lead to shorter realized generation intervals.
However, this explanation is also less likely given similarities in the inferred incubation-period distributions.

Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections. 
To better illustrate the network effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days;
in this case, both variants have identical intrinsic generation-interval distributions.
However, if both individuals were each living in a household consisting of four other people, an individual infected with variant A will infect one household member everyday; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
Previous simulations showed that such network effects can have considerable impact on realized generation intervals even during the initial exponential growth phase, when susceptible depletion is negligible at the population level \citep{park2020inferring}.
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).

Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but it is still uncertain how infectiousness profiles differ intrinsically between Omicron and Delta.
In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
and can therefore be expected to improve estimates of effective reproduction numbers.

Our study also has important implications for estimating transmission advantages of new SARS-CoV-2 variants.
In the example we consider, neglecting differences in the generation-interval distributions leads to $\approx 10\%$ bias in the estimates of the reproduction advantage (i.e., the ratio between the reproduction numbers of the Omicron and Delta variants).
More generally, the bias in inferring the reproduction advantage of an emerging variant is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the previously assumed generation-interval distributions and estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.

>>>>

Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
\cite{ito2022estimating,selby2022generation} estimated the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
The realized generation intervals represent time between actual infection events, and are different from the intrinsic generation intervals, which reflect the average profile of infectiousness of infected individuals over the course of their infections \citep{champredon2015intrinsic}.

Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals.
For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} may lead to faster recovery, which could in turn shorten the intrinsic generation interval \citep{roberts2007model}.
Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals).
In our analysis, a greater proportion of Omicron than Delta infections would have been in people with pre-existing immunity (because these people are less likely to be infected by Delta);
this allowed the Omicron variant to rapidly replace the previously dominant Delta variant at the population level.
However, the individual-level relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is difficult to predict given its high immune evasiveness:
some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasiveness may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, are separate from possible changes in the generation-interval distribution that we have outlined above.
Even if Delta and Omicron variants had identical intrinsic generation-interval distributions, immune evasiveness alone would allow the Omicron variant to become dominant.
Therefore, we might expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections to be more strongly influenced by the intrinsic differences in their viral kinetics.

Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions, their realized generation-interval distributions could still differ.
For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, could also lead to shorter realized generation intervals.
However, this explanation is also less likely given similarities in the inferred incubation-period distributions.

Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized. 
To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simplest example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).

Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but the difference between the intrinsic infectiousness profiles remains unclear.
In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
and can therefore be expected to improve estimates of effective reproduction numbers.

Our study also has important implications for estimating transmission advantages of new SARS-CoV-2 variants.
In the example we consider, neglecting differences in the generation-interval distributions leads to $\approx 10\%$ bias in the estimates of the reproduction advantage (i.e., the ratio between the reproduction numbers of the Omicron and Delta variants).
More generally, the bias in inferring the reproduction advantage of an emerging variant is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province.
